---
title: "Porto Alegre Accidents"
author: "Felipe Paula"
date: "31 de outubro de 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Accidents in the city of Porto Alegre, Rio Grande do Sul, Brazil

In this document, we analyze the data regarding the accidents that occured in the city of POA (Porto Alegre). The data is provided by the city's center of open data #DataPOA. 

```{r}
file.name <- "acidentes-2016.csv"
if(!file.exists(file.name)){
  download.file("http://datapoa.com.br/storage/f/2017-08-03T13%3A19%3A45.538Z/acidentes-2016.csv",
	destfile=file.name)
}
df <- read.csv2(file.name, stringsAsFactors = FALSE)

```

There's also a dictionary of the semantics of the data available in the website.

```{r}
file.name <- "dicionariodedadosacidentes.pdf"
if(!file.exists(file.name)){
  download.file("http://datapoa.com.br/storage/f/2017-05-19T19%3A49%3A51.583Z/dicionariodedadosacidentes.pdf",
	destfile=file.name)
}
```

## Is there a time of the year with more accidents?

For answer this question we examine in different levels of granularity, days of week and months. For days of week:

```{r}
require(tidyverse)

df %>%
  group_by(DIA_SEM) %>%
  summarise(n_accidents = n())
```

It seems that the number of accidents in the weekend (SABADO and DOMINGO) is lower than during the week. One hypothesis is that there is few cars on the streets. _Fridays present higher absolute number of accidents_. Next analysis should include discrimination by the type of accident.

```{r}
df %>%
  group_by(DIA_SEM) %>%
  summarise(n_accidents = n())
```


For months:

```{r}
df %>%
  group_by(MES) %>%
  summarise(n_accidents = n())

df %>%
  group_by(MES) %>%
  summarise(n_accidents = n()) %>%
  ggplot(aes(x=MES, y=n_accidents, group=1)) +
  geom_bar(stat="identity")
```


February present less accidents, however it have fewer days. _We should normalize by the number of days_.

For days:

```{r}
df %>%
  group_by(TIPO_ACID,DATA) %>%
  summarise(n_accidents = n())  %>%
  ggplot(aes(x=DATA, y=n_accidents, color=TIPO_ACID,group = 1)) +
  facet_grid(TIPO_ACID ~ .) + 
  geom_line()
```


```{r}
require(lubridate)
df %>%
  mutate(DATA = as.Date(DATA)) %>%
  
  mutate(WEEK = week(DATA)) -> df

df %>%
  group_by(WEEK) %>%
  summarise(n_accidents = n()) %>%
  filter(WEEK != 1, WEEK != 53) -> week.accidents

week.accidents %>%
  summarise(mean_acc = mean(n_accidents),sd_acc = sd(n_accidents)) -> week.acc.stats

week.accidents %>%
  ggplot(aes(x=WEEK, y=n_accidents,group = 1)) +
  annotate("rect", xmin = 1, xmax = 5, ymin = 0, ymax = 300,
  alpha = 0, colour = 'black',linetype='dashed') + 
  annotate("text", x = 3, y = 45, label = "JAN") + 
  annotate("text", x = 3, y = 25, label = "1021") + 

  annotate("rect", xmin = 10, xmax = 14, ymin = 0, ymax = 300,
  alpha = 0, colour = 'black',linetype='dashed') +
  annotate("text", x = 12, y = 45, label = "MAR") + 
  annotate("text", x = 12, y = 25, label = "1172") + 

  annotate("rect", xmin = 22, xmax = 26, ymin = 0, ymax = 300,
  alpha = 0, colour = 'black',linetype='dashed') + 
  annotate("text", x = 24, y = 45, label = "JUN") +
  annotate("text", x = 24, y = 25, label = "1124") +

  annotate("rect", xmin = 31, xmax = 35, ymin = 0, ymax = 300,
  alpha = 0, colour = 'black',linetype='dashed') + 
  annotate("text", x = 33, y = 45, label = "AGO") + 
  annotate("text", x = 33, y = 25, label = "1098") + 

  annotate("rect", xmin = 39, xmax = 44, ymin = 0, ymax = 300,
  alpha = 0, colour = 'black',linetype='dashed') + 
  annotate("text", x = 41.5, y = 45, label = "OCT") + 
  annotate("text", x = 41.5, y = 25, label = "1041") + 

  annotate("rect", xmin = 48, xmax = 52, ymin = 0, ymax = 300,
  alpha = 0, colour = 'black',linetype='dashed') +
  annotate("text", x = 50, y = 45, label = "DEC") +
  annotate("text", x = 50, y = 25, label = "1035") +

  annotate("segment", x = -Inf, xend = Inf, y = 242, yend = 242) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 242 - 31.5, ymax = 242 + 31.5,
  alpha = 0.2, colour = 'orange', fill='orange') +
  annotate("text", x = 57, y = 242, label = "Mean \n 242 (31.5)") +
  annotate("text", x = 60, y = 242, label = "") +
  scale_x_continuous(breaks=seq(0,53,5)) + 
  geom_line() +
  ylab('Number of accidents \n') + 
  xlab('Week') +
  ggtitle('Number of accidents per week') + 

  theme_minimal() + 
  theme(legend.position="none",
        axis.text.x = element_text(angle = 25, hjust = .5,size = 12),
        axis.title =element_text(size = 15),
        axis.text.y = element_text(size = 12))

```


```{r}

df %>%
  mutate(MES = as.factor(MES)) %>%
  group_by(DATA,MES) %>%
  summarise(n_accidents = n()) -> accidents.per.date

summary(aov(n_accidents ~ MES, accidents.per.date))


```

The time series seem periodic. I'd like to know how associated are the accident types.

```{r}
df %>%
  group_by(TIPO_ACID,DATA) %>%
  summarise(n_accidents = n()) %>%
  spread(TIPO_ACID,n_accidents) %>%
  replace(is.na(.),0) %>%
  select(-DATA, -EVENTUAL, -INCENDIO, -QUEDA, -TOMBAMENTO) %>%
  as.matrix() %>%
  #acf() ##cross correlation plots
  cor(method='pearson') %>%
  as.tibble() %>%
  mutate(id2 = colnames(.)) %>%
  gather(id,value, 1:5) %>%
  ggplot(aes(x=id, y=id2, fill=value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white",
   midpoint = 0, limit = c(-1,1), space = "Lab",
    name="Pearson\nCorrelation") +
  theme_minimal()

```

They are kind of associated. This test did not gave me any insight.


```{r}
df %>%
  #filter(TIPO_ACID != c('EVENTUAL', 'INCENDIO', 'QUEDA', 'TOMBAMENTO')) %>%
  group_by(DATA) %>%
  summarise(n_accidents = n()) -> acc.ts

SMA(acc.ts$n_accidents,14) %>%
plot(.,type='l')
```

There's a peak around end of march. Maybe it's the easter's eve. It's not during the summer break and it's a big holyday.  

```{r}
## Takes a day in portuguese and translates to english
translate.week.days <- function(weekday){
  
  dictionary <- list("SEGUNDA-FEIRA" = 'Monday',
                     "TERCA-FEIRA" = 'Tuesday',
                     "QUARTA-FEIRA" = "Wednesday",
                     "QUINTA-FEIRA"='Thursday',
                     "SEXTA-FEIRA"='Friday',
                     "SABADO" = 'Saturday',
                     "DOMINGO"='Sunday')
  return(unlist(dictionary[weekday]))
}

df %>%
  group_by(DIA_SEM,DATA) %>%
  summarise(n_accidents = n()) %>%
  ungroup() %>%
  mutate(DIA_SEM = translate.week.days(DIA_SEM)) -> accidents.per.week

# accidents.per.week %>%
#   mutate(DIA_SEMS = str_to_lower(DIA_SEM)) -> accidents.per.week

accidents.per.week$DIA_SEM <- ordered(factor(accidents.per.week$DIA_SEM),
                                      levels = c("Monday", "Tuesday", "Wednesday","Thursday", "Friday", "Saturday", "Sunday"))

require(RColorBrewer)

accidents.per.week %>%
  ggplot(aes(x=DIA_SEM, y=n_accidents, fill = DIA_SEM)) +
  stat_boxplot(geom = "errorbar", width = 0.5,size=0.8) +
  geom_boxplot(width=0.5) +
  scale_fill_brewer(palette="Oranges") + 
  scale_y_continuous(limits = c(0, 70)) +
  ylab('Number of accidents\n')+
  theme_minimal() + 
  theme(legend.position="none",
        axis.text.x = element_text(angle = 25, hjust = .5,size = 12),
        axis.title.x=element_blank(),
        axis.title.y=element_text(size = 15),
        axis.text.y = element_text(size = 12))
```


```{r}

accidents.per.week$DIA_SEM <- factor(accidents.per.week$DIA_SEM)

accidents.per.week %>%
group_by(DIA_SEM) %>%
summarise(sds = sd(n_accidents))

accidents.per.week.anova <- aov(n_accidents ~ DIA_SEM,data=accidents.per.week)
# summary(accidents.per.week.anova)

TukeyHSD(accidents.per.week.anova)
```

2016 is a leap year.



## How many vehicles are usually involved?

```{r}
df %>%
  group_by(TIPO_ACID) %>%
  summarise(auto = sum(AUTO), 
            taxi = sum(TAXI), 
            lotacao = sum(LOTACAO), 
            oni_urbanos = sum(ONIBUS_URB), 
            oni_interior = sum(ONIBUS_INT),
            caminhao = sum(CAMINHAO),
            moto = sum(MOTO),
            carroca = sum(CARROCA),
            bicicleta = sum(BICICLETA),
            outro = sum(OUTRO))
```


```{r}
df %>%
  mutate(tv = AUTO + TAXI + LOTACAO + ONIBUS_URB + ONIBUS_INT + ONIBUS_MET + CAMINHAO + MOTO + CARROCA + BICICLETA + OUTRO)  -> total.vehicles

total.vehicles$tv <- as.factor(total.vehicles$tv)

total.vehicles %>%
ggplot(aes(tv)) +
ylab('Number of accidents\n')+
xlab('\nNumber of vehicles involved')+
# geom_histogram(stat="count",binwidth=.5, colour="black", fill="gold2") + 
geom_bar(stat = "count", colour="black", fill="gold2")+
geom_text(stat='count',aes(label=..count..),vjust=-1)+
ylim(0, 12000) + 
theme_minimal() +
  theme(
        axis.text.x = element_text(size = 12),
        axis.title=element_text(size = 15),
        axis.text.y = element_text(size = 12))

```

```{r}

total.vehicles %>%
  filter(tv == 0)

```



